Active Learning for SAT Solver Benchmarking {\large (extended and revised version)}

### 2024-05-25 paper_jar/paper.tex
Review changes in non-extended parts of JAR paper
I.e., excluding the two subsection containing new material. Revert or fix
some changes.

\abstract{
Benchmarking is a crucial phase in the development of algorithms.
This also applies to SAT solvers for the problem of propositional satisfiability.
Benchmark selection is about choosing representative problem instances that reliably discriminate solvers based on their runtime.
In this paper, we present a dynamic benchmark selection approach based on active learning.
Our approach predicts the rank of a new solver among its competitors with minimum runtime and maximum rank prediction accuracy.
This approach was evaluated on the Anniversary track dataset from the 2022 SAT Competition.
Our selection approach can predict the rank of a new solver after approximately \SI{10}{\%} of the time it would take to run the solver on all instances of this dataset, with a prediction accuracy of approximately \SI{92}{\%}. 
Additionally, the importance of instance families in the selection process is discussed.  
In conclusion, our tool offers a reliable method for solver engineers to assess a new solver's performance in an efficient manner.
}

@@

\section{Introduction}
\label{sec:intro}
% TODO: if appropriate, extend with motivation/results regarding runtime discretization and prediction models

One of the main phases of algorithm engineering is benchmarking.
This also applies to propositional satisfiability (SAT), the canonical $\mathcal{NP}$-complete problem.
Benchmarking is, however, quite expensive regarding the runtime of experiments.
While it is still feasible to benchmark a single or fixed amount of SAT solvers, developing new, competitive SAT solvers requires extensive experimentation with a variety of ideas, which often results in a combinatorial explosion of the configuration space of each algorithm~\cite{HutterHL11}.
In particular, a new solver idea is rarely best on the first try.
Thus, it is highly desirable to reduce benchmarking time and discard unpromising ideas early, allowing to test more approaches or spend more time on promising ones.
The field of SAT solver benchmarking is well established, yet traditional benchmark selection approaches do not optimize benchmark runtime.
Instead, they focus on selecting a representative set of instances for ranking solvers~\cite{Gelder11,HoosKSS13}.
With regard to the ranking of solvers, SAT Competitions typically employ the \mbox{PAR-2} score, which, given a time limit of $\tau$, is the average runtime with a penalty of $2 \tau$ for timeouts~\cite{FroleyksHIJS21}.

@@

The approach presented in this paper meets several criteria for benchmarking that are considered desirable.
Rather than outputting a binary classification, namely whether the new solver is worse than an existing solver or not, we provide a scoring function that shows by which margin a solver is worse and how similar it is to existing solvers.
In particular, our approach enables ranking the new solver amidst a set of existing solvers.
We show that for ranking, it is not necessary to predict the exact per-instance runtime of the solvers, which would be a more challenging task.
Furthermore, we optimize the \emph{runtime} required for our strategy to arrive at its conclusion.
To this end, we use instance \emph{features} and known solver \emph{runtimes}.
Moreover, we select instances \emph{non-randomly} and \emph{incrementally}.
In particular, we consider runtime information from already done experiments when choosing the next.
By doing so, we can control the properties of the benchmarking approach, such as its required runtime.
Our approach is \emph{scalable} in that it ranks a new solver $\hat{a}$ among any number of known solvers $\mathcal{A}$.
In particular, we only subsample the benchmark once instead of comparing pairwise against each other solver~\cite{MatriconAFSH21}.

@@

This article is an extended and refined version of our conference paper with the title ``Active Learning for SAT Solver Benchmarking''~\cite{fuchs2023active}.
In particular, we significantly extended the analysis of runtime-prediction approaches, which led to our choice of a solver runtime model for use in benchmark instance selection (cf.~Section~\ref{sec:exdesign:disc-pred}).
Furthermore, we expanded the evaluation of which benchmark instances are selected by our approach (cf.~Section~\ref{sec:eval:instance}).
% TODO: before submission, we should check which other changes we made and list them here accordingly

@@

Benchmarking is of great interest in numerous research areas and represents an active field of research in its own right.
Studies have shown that the compilation of benchmark instances poses a number of challenges.
The use of biased benchmarks can easily lead to fallacious interpretations~\cite{abs-2107-07002}.
Benchmarking also involves a number of interchangeable elements, including the performance measures used, the way the measures are aggregated, and the way missing values are handled.
It is possible that questionable research practices could alter these elements a-posteriori in order to meet expectations, thereby skewing the results~\cite{NiesslHWCB22}.

@@

Benchmark selection is essential for competitions such as the SAT Competition.
In such competitions, the organizers define the rules for composing the benchmarks.
These selection strategies are primarily static, in that they do not depend on particular solvers to distinguish.
Balint et al. provide an overview of benchmark-selection criteria in different solver competitions~\cite{balint2015overview}.
Froleyks et al. describe benchmark selection in recent SAT~competitions~\cite{FroleyksHIJS21}.
Manthey and Möhle propose a feature-based approach to remove redundancy from competition benchmarks~\cite{manthey2016better}, while M{\i}s{\i}r presents a similar feature-based approach to reduce benchmarks by matrix factorization and clustering~\cite{misir2021benchmark}.

In their study, Hoos et al.~\cite{HoosKSS13} discuss the desirable properties of SAT benchmark instances.
They identify three key selection criteria: instance variety to avoid over-fitting, adapted instance hardness (not too easy but also not too hard), and the avoidance of duplicate instances. 
To filter instances that are too similar, they employ a distance-based approach with the SATzilla features~\cite{XuHHL08,features}.
It should be noted, however, that the approach does not optimize for benchmark runtime.
Instead, instances are selected at random, with the exception of constraints on instance hardness and feature distance.

@@

Matricon~et~al. present an incremental benchmark selection approach~\cite{MatriconAFSH21}.
Their \emph{per-set efficient algorithm selection problem} (PSEAS) is similar to our \emph{New-Solver Problem}, as given in Definition~\ref{def:new-solver-problem}.
To differentiate between \emph{a pair of} SAT solvers, the authors propose an iterative approach to select a solver-specific subset of instances until a desired confidence level is reached.
This is achieved by calculating a scoring metric for all unselected instances, running the experiment with the highest score, and updating the confidence.
Their approach ticks off most of the desired features in Table~\ref{tab:requirements}.
However, the approach only compares solvers binarily rather than providing a \emph{scoring}.
Consequently, it is unclear how similar two given solvers are or on which instances they behave similarly.
Furthermore, a significant shortcoming is the lack of \emph{scalability} with the number of solvers.
Comparing only pairs of solvers, evaluating a new solver requires sampling a separate benchmark for each existing solver.
In contrast, our approach allows comparing a new solver against a set of existing solvers by sampling only one benchmark.
In contrast to our approach, which allows for the comparison of a new solver against a set of existing solvers by sampling only one benchmark, their pairwise method requires the evaluation of a new solver to be conducted on a separate benchmark for each existing solver.

@@

In passive machine learning, prediction models are trained on datasets with given instance labels (cf.~Fig.~\ref{fig:passive}).
In contrast, active learning (AL) begins with no or little labeled data.
AL then repeatedly selects interesting problem instances for which to acquire labels, gradually improving the prediction model (cf.~Fig.~\ref{fig:activepool}).
AL methods are particularly advantageous when acquiring labels is computationally expensive, such as obtaining solver runtimes.
Without AL methods, it is not evident which instances to label and which not.
The objective is to maximize the utility of instances to our model, which is measured by the rank prediction accuracy.
Concurrently, the objective is to minimize the cost of label acquisition, which is measured by the instance's predicted runtime.
Consequently, the objective is to develop an accurate prediction model without the necessity of labeling each data point.

@@

Algorithm~\ref{algALBenchmark} outlines our benchmarking framework. 
Given a set of solvers~$\mathcal{A}$, instances~$\mathcal{I}$ and runtimes~$r$, we first initialize a prediction model~$\mathcal{M}$ for the new solver $\hat a \not\in \mathcal{A}$ (Line~1).
The prediction model~$\mathcal{M}$ is used to select instances repeatedly (Line~4) for benchmarking the new solver $\hat a$ (Line~5). 
The acquired result is subsequently employed to update the prediction model, denoted by~$\mathcal{M}$ (Line~7). 
When the stopping criterion is met (Line~3), the benchmarking loop is terminated and the final score of $\hat{a}$ is estimated (Line~8). 
This score, along with the acquired instances and runtime measurements, is returned by Algorithm~\ref{algALBenchmark}.

@@

The model $M$ provides a runtime-label prediction function $f : \mathcal{\hat A} \times \mathcal{I} \rightarrow \mathbb{R}$ for all solvers $\mathcal{\hat A} := \mathcal{A} \cup \lbrace \hat a \rbrace$.
This prediction function powers instance selection as described in Section~\ref{sec:main:selection}.
During model updates (Algorithm~\ref{algALBenchmark},~Line~7), the prediction function $f$ is trained to predict a transformed version of the acquired runtime data set~$\mathcal{R}$.
The transformation of the runtime data set is described in the subsequent section.
The features described in Section~\ref{sec:exdesign:data} serve as the input to the model.
Furthermore, it should be noted that a new prediction model is constructed in each iteration, as the runtime of the experiments (Line~5) dominates the training time of the model by a significant margin.
Finally, the score of the new solver, denoted by $\hat a$, is predicted with the prediction function~$f$ (Line~8).

@@

\paragraph{Uncertainty Sampling}
The uncertainty sampling strategy selects the instance that is closest to the model's decision boundary.
This is achieved by selecting the instance $e \in \mathcal{I} \setminus \tilde{\mathcal{I}}$ that minimizes $U(e)$, which is specified by the following equation.
%
\begin{equation*}
  \operatorname{U}(e) := \left\lvert \frac{1}{k} - \max_{n \in \left\lbrace 1, \dots, k \right\rbrace} f'\!\left(\hat{a}, e\right)_{n} \right\rvert
\end{equation*}

\paragraph{Information-Gain Sampling}
The information-gain sampling strategy selects the instance with the highest expected entropy reduction regarding the runtime labels of the instance.
To be more precise, we select the instance $e \in \mathcal{I} \setminus \tilde{\mathcal{I}}$ that maximizes $IG(e)$, which is specified in the following equation.
%
\begin{equation*}
  \operatorname{IG}(e) := \operatorname{H}(e) - \sum_{n = 1}^{k} f'(\hat{a}, e)_{n} \operatorname{\hat H}_n(e)
\end{equation*}
%
In the equation, $\operatorname{H}(e)$ denotes the entropy of the runtime labels $\gamma(a, e)$ over all $a \in \mathcal{A}$ and $\operatorname{H}(e, n)$ denotes the entropy of these labels plus $n$ as the runtime label for $\hat{a}$.
The term $\operatorname{\hat H}_n(e)$ is computed for every possible runtime label $n \in \{1, \dots, k\}$.
By maximizing the information gain, instances are selected that identify solvers with similar behavior.

@@

\paragraph{Wilcoxon Stopping Criterion}
The Wilcoxon stopping criterion is employed to halt the active-learning process based on the confidence that the predicted runtime labels of the new solver are sufficiently different from those of existing solvers.
This criterion is loosely inspired by Matricon et.~al.~\cite{MatriconAFSH21}.
The average $p$-value $W_{\hat{a}}$ of a Wilcoxon signed-rank test $\operatorname{w}(S,P)$, as given by the following equation, of the two runtime label distributions ${S=\{ \gamma(a, e) \mid e \in \mathcal{I} \}}$ for an existing solver $a$ and \mbox{$P=\{ f(\hat a, e) \mid e \in \mathcal{I} \}$} of the new solver $\hat{a}$ is used to assess the statistical significance of the predicted runtime labels.
%
\begin{equation*}
  W_{\hat{a}} := \frac{1}{\lvert \mathcal{A} \rvert} \sum_{a \in \mathcal{A}} \operatorname{w}(S, P)
\end{equation*}
%
In order to improve the stability of this criterion, an exponential moving average is employed to mitigate the impact of outliers, with the process terminated when the value of $W^{(i)}_{\exp}$ drops below a fixed threshold.
%
\begin{align*}
  W_{\exp}^{\left(0\right)} &:= 1\\
  W_{\exp}^{\left(i\right)} &:= \beta W_{\hat{a}} + \left(1 - \beta\right) W_{\exp}^{\left(i - 1\right)}
\end{align*}

\paragraph{Ranking Stopping Criterion}
In contrast, the stopping criterion for the ranking process is less sophisticated than the previous one.
In this case, the active learning process is terminated if the ranking induced by the model's predictions (Equation~\ref{eq:rankingeq}) remains unchanged within the last~$l$~iterations.
However, it should be noted that the concrete values of the predicted score $s_{\hat a}$ may still change.
Nevertheless, in this case, the induced ranking is of sole interest.

@@

In light of the preceding instantiations of Algorithm~\ref{algALBenchmark}, this section outlines the experimental design employed in the presented study.
This includes an overview of the evaluation framework, the data sets employed, the hyperparameter selections, and the implementation details.

@@

As stated in the Introduction, this work addresses the \emph{New-Solver Problem} (cf.~Definition~\ref{def:new-solver-problem}).
As described in Section~\ref{sec:main:model}, a prediction model provides us with an estimated scoring value for the new solver~$\hat{a}$.
%
To evaluate the efficacy of a concrete instantiation of Algorithm~\ref{algALBenchmark}, which entails a specific choice for all subroutines, we perform cross-validation on our set of solvers.

This is demonstrated in Algorithm~\ref{alg:eval}.
Each solver assumes the role of the new solver~$\hat{a}$ once (Line~2) and is consequently excluded from the set of solvers $\mathcal{A}$ (Line~3).
After running our active-learning framework for solver~$\hat{a}$ (Line~4), we compute the value of both our optimization goals, i.e., ranking accuracy and runtime.
%
The \emph{ranking accuracy}, expressed in the interval $[0,1]$ (higher is better), is defined as the fraction of correctly ranked pairs of elements, namely, $\left(\hat{a}, a\right)$, where $a$ belongs to the set of all solvers $\mathcal{A}$ ranked by the ground-truth scoring function $\operatorname{par_2}$ (Lines~5-8).
%
The \emph{fraction of runtime} required by the algorithm to reach its conclusion is represented by $O_{\operatorname{rt}}$, which lies within the interval $[0, 1]$ (lower is better).
This metric compares the accumulated runtimes over the sampled instances to the accumulated runtimes over all instances in the dataset (Lines~9-13).
Once all cross-validation results have been collected (Line~14), the output metrics are averaged (Line~15).

The procedure is designed to identify an approach that maximizes Equation~\eqref{eq:opt}, in which $\delta \in \left[0, 1\right]$ is used for linear weighting, with weighting factor $\delta \in \left[0, 1\right]$, between the two optimization goals $O_{\operatorname{acc}}$ and $O_{\operatorname{rt}}$.
The plotting of the approaches that maximize $O_\delta$ for all $\delta \in \left[0, 1\right]$ on an $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagram provides us with a Pareto front of the best approaches for different optimization-goal weightings.
%
\begin{equation}
  O_\delta := \delta O_{\operatorname{acc}} + \left(1 - \delta\right) \left(1 - O_{\operatorname{rt}}\right) \enspace \textrm{,}
  \label{eq:opt}
\end{equation} 
%
%
\subsection{Data}
\label{sec:exdesign:data}

In our experiments we work with the dataset of the SAT~Competition~2022 Anniversary Track~\cite{sat2022}.
The dataset consists of 5355 instances with corresponding runtime data of 28 sequential SAT solvers.
We also use a database of 56 instance features\footnote{\url{https://benchmark-database.de/getdatabase/base_db}} from the Global Benchmark Database~(GBD) by Iser~et~al.~\cite{IserS18}.
They include instance size features and node distribution statistics for several graph representations of SAT instances, and are primarily inspired by the SATzilla~2012 features described in~\cite{features}.
All features are numeric and have no missing values.
We drop 10 of the 56 features due to zero variance.
In total, the prediction models have access to 46 instance features and 27 out of 28 runtime features, excluding the respective new solver~$\hat{a}$.

@@

\subparagraph{Ranking}
Regarding \emph{ranking}, we experiment with the following instantiations.

@@

\subparagraph{Selection}
We experiment with the following instantiations of instance \emph{selection}.
We experiment with the following instantiations of the instance \emph{selection} criteria.
Since the potential runtime of the experiments is orders of magnitude larger than the model update time, we consider incrementing our benchmark one instance at a time rather than using batches, which is also proposed in recent advances in active learning~\cite{SinhaED19,2019gaal}.
A drawback of this is the lack of parallel execution of runtime experiments.

@@

\subparagraph{Stopping}
We evaluate the following \emph{stopping} criteria.

@@

\subsection{Runtime Discretization and Prediction}
\label{sec:exdesign:disc-pred}
One of the most important parts of our framework is the choice of solver runtime model to inform active learning selection decisions.
In this section, we detail how clustering solver runtimes can lead to nuanced runtime labels, and which machine learning models are best at predicting them.

@@

% MI: table with cluster prediction performance: (merged with existing text)
Since runtime prediction is quite challenging, we discretize the runtimes.
To define prediction targets for each instance, we cluster the solver runtimes into $k$ clusters.
We denote this adapted solver scoring function which assigns a solver to its cluster for each instance with $s_k$ as defined in Equation~\eqref{eq:rankingeq}.
%
In preliminary experiments, the most \emph{useful} labels were produced with hierarchical clustering and a log-single-link criterion.
In our chosen hierarchical procedure, each non-timeout runtime starts in a separate interval.
We then gradually merge intervals with the smallest single-link logarithmic distance until the desired number of partitions is reached.
The clustering is instance-specific, i.e., the bounds are chosen to produce groups of solvers with similar performance for each instance.
Other clustering approaches we have tried include hierarchical clustering with mean, median, and full-link criteria, as well as $k$-means and spectral clustering.

@@

Our final choice for the prediction model is the stacking ensemble of a quadratic discriminant analysis and a random forest.
Both of these models can learn non-linear relationships between the instance features and the runtime labels.
Figure~\ref{fig:confusion-matrices} shows the confusion matrices of the stacking ensemble for predicting timeouts and $C_3$ clusters.
Both perform reasonably well, with the $C_3$ cluster prediction model performing slightly worse than the timeout prediction model.
Notably, for the $C_3$ prediction model, there is less confusion between the non-timeout labels (labels 1 and 2) and the timeout label (label 3) than between the non-timeout labels (labels 1 and 2).
This suggests that deciding on timeouts is easier than distinguishing between different groups of solvers.

>>>>

\abstract{
Benchmarking is a crucial phase in the development of algorithms.
This also applies to solvers for the propositional satisfiability (SAT) problem.
Benchmark selection is about choosing representative problem instances that reliably discriminate solvers based on their runtime.
In this paper, we present a dynamic benchmark selection approach based on active learning.
Our approach predicts the rank of a new solver among its competitors with minimum runtime and maximum rank prediction accuracy.
We evaluated this approach on the Anniversary Track dataset from the SAT Competition 2022.
Our selection approach can predict the rank of a new solver after approximately \SI{10}{\%} of the time it would take to run the solver on all instances of this dataset, with a prediction accuracy of approximately \SI{92}{\%}. 
Additionally, we discuss the importance of instance families in the selection process.
In conclusion, our tool offers a reliable method for solver engineers to assess a new solver's performance efficiently.
}

@@

\section{Introduction}
\label{sec:intro}

One of the main phases of algorithm engineering is benchmarking.
This also applies to propositional satisfiability (SAT), the canonical $\mathcal{NP}$-complete problem.
Benchmarking is, however, quite expensive regarding the runtime of experiments.
While it is still feasible to benchmark a single or small number of SAT solvers, developing new, competitive SAT solvers requires extensive experimentation with a variety of ideas.
The latter often results in a combinatorial explosion of the configuration space of each algorithm~\cite{HutterHL11}.
In particular, a new solver idea is rarely best on the first try.
Thus, it is highly desirable to reduce benchmarking time and discard unpromising ideas early, allowing to test more approaches or spend more time on promising ones.
The field of SAT solver benchmarking is well established, yet traditional benchmark selection approaches do not optimize benchmark runtime.
Instead, they focus on selecting a representative set of instances for ranking solvers~\cite{Gelder11,HoosKSS13}.
In this regard, SAT Competitions typically employ the \mbox{PAR-2} score, which, given a time limit of $\tau$, is the average runtime with a penalty of $2 \tau$ for timeouts~\cite{FroleyksHIJS21}.

@@

The approach presented in this paper meets several criteria for benchmarking that are considered desirable (cf.~Table~\ref{tab:requirements}).
Rather than outputting a binary classification, namely whether the new solver is worse than an existing solver or not, we provide a \emph{scoring} function that shows by which margin a solver is worse and how similar it is to existing solvers.
In particular, our approach enables \emph{ranking} the new solver amidst a set of existing solvers.
We show that for ranking, it is not necessary to predict the exact per-instance runtime of the solvers, which would be a more challenging task.
Furthermore, we optimize the \emph{runtime} required for our strategy to arrive at its conclusion.
To this end, we use instance \emph{features} and known solver \emph{runtimes}.
Moreover, we select instances \emph{non-randomly} and \emph{incrementally}.
In particular, we consider runtime information from already done experiments when choosing the next.
By doing so, we can control the properties of the benchmarking approach, such as its required runtime.
Our approach is \emph{scalable} in that it ranks a new solver $\hat{a}$ among any number of known solvers $\mathcal{A}$.
In particular, we only subsample the benchmark once instead of comparing pairwise against each other solver~\cite{MatriconAFSH21}.

@@

This article is an extended and refined version of our conference paper with the title ``Active Learning for SAT Solver Benchmarking''~\cite{fuchs2023active}.
In particular, we significantly extended the analysis of runtime-prediction approaches, which determined our chosen solver runtime model for instance selection (cf.~Section~\ref{sec:exdesign:disc-pred}).
Furthermore, we expanded the evaluation of which benchmark instances are selected by our approach (cf.~Section~\ref{sec:eval:instance}).

@@

Benchmarking is of great interest in numerous research areas and represents an active field of research in its own right.
Studies have shown that the compilation of benchmark instances poses a number of challenges.
The use of biased benchmarks can easily lead to fallacious interpretations~\cite{abs-2107-07002}.
Benchmarking also involves a number of interchangeable elements, including the performance measures used, the way the measures are aggregated, and the way missing values are handled.
Questionable research practices could alter these elements a-posteriori in order to meet expectations, thereby skewing the results~\cite{NiesslHWCB22}.

@@

Benchmark selection is essential for competitions such as the SAT Competition.
In such competitions, the organizers define the rules for composing the benchmarks.
These selection strategies are primarily static, in that they do not depend on particular solvers to distinguish.
Balint et al. provide an overview of benchmark-selection criteria in different solver competitions~\cite{balint2015overview}.
Froleyks et al. describe benchmark selection in recent SAT~competitions~\cite{FroleyksHIJS21}.
Manthey and Möhle propose an approach to remove redundancy from competition benchmarks~\cite{manthey2016better}, considering feature equivalence of formulas.
M{\i}s{\i}r also presents a feature-based approach to reduce benchmarks, using matrix factorization and clustering~\cite{misir2021benchmark}.

Hoos et al.~\cite{HoosKSS13} discuss desirable properties of SAT benchmark instances.
They identify three key selection criteria: instance variety to avoid over-fitting, adapted instance hardness (not too easy but also not too hard), and the avoidance of duplicate instances. 
To filter instances that are too similar, they employ a distance-based approach with the SATzilla features~\cite{XuHHL08,features}.
It should be noted, however, that the approach does not optimize for benchmark \emph{runtime}.
Instead, instances are selected \emph{at random}, with the exception of constraints on instance hardness and feature distance.

@@

Matricon~et~al. present an incremental benchmark selection approach~\cite{MatriconAFSH21}.
Their \emph{per-set efficient algorithm selection problem} (PSEAS) is similar to our \emph{New-Solver Problem}, as given in Definition~\ref{def:new-solver-problem}.
To differentiate between \emph{a pair of} SAT solvers, the authors propose an iterative approach to select a solver-specific subset of instances until a desired confidence level is reached.
This is achieved by calculating a scoring metric for all unselected instances, running the experiment with the highest score, and updating the confidence.
Their approach ticks off most of the desired features in Table~\ref{tab:requirements}.
However, the approach only compares solvers binarily rather than providing a \emph{scoring}.
Consequently, it is unclear how similar two given solvers are or on which instances they behave similarly.
Furthermore, a significant shortcoming is the lack of \emph{scalability} with the number of solvers.
Comparing only pairs of solvers, evaluating a new solver with their method requires sampling a separate benchmark for each existing solver.
In contrast, our approach allows comparing a new solver against a set of existing solvers by sampling only one benchmark.

@@

In passive machine learning, prediction models are trained on datasets with given instance labels (cf.~Fig.~\ref{fig:passive}).
In contrast, active learning (AL) begins with no or little labeled data.
AL then repeatedly selects interesting problem instances for which to acquire labels, gradually improving the prediction model (cf.~Fig.~\ref{fig:activepool}).
AL methods are particularly advantageous when acquiring labels is computationally expensive, such as obtaining solver runtimes.
Without AL methods, it is not evident which instances to label and which not.
The first goal is to maximize the utility of instances to our model, which is measured by the rank prediction accuracy.
The second goal is to minimize the cost of label acquisition, which is measured by the instance's predicted runtime.
Consequently, the overall objective is to develop an accurate prediction model without the necessity of labeling each data point.

@@

Algorithm~\ref{algALBenchmark} outlines our benchmarking framework. 
Given a set of solvers~$\mathcal{A}$, instances~$\mathcal{I}$ and runtimes~$r$, we first initialize a prediction model~$\mathcal{M}$ for the new solver $\hat a \not\in \mathcal{A}$ (Line~1).
The prediction model~$\mathcal{M}$ is used to select instances repeatedly (Line~4) for benchmarking the new solver $\hat a$ (Line~5). 
The acquired result is subsequently employed to update the prediction model, denoted by~$\mathcal{M}$ (Line~7). 
When the stopping criterion is met (Line~3), the benchmarking loop is terminated and the final score of $\hat{a}$ is estimated (Line~8). 
Algorithm~\ref{algALBenchmark} returns this score, along with the acquired instances and runtime measurements.

@@

The model $M$ provides a runtime-label prediction function $f : \mathcal{\hat A} \times \mathcal{I} \rightarrow \mathbb{R}$ for all solvers $\mathcal{\hat A} := \mathcal{A} \cup \lbrace \hat a \rbrace$.
This prediction function powers instance selection as described in Section~\ref{sec:main:selection}.
During model updates (Algorithm~\ref{algALBenchmark},~Line~7), the prediction function $f$ is trained to predict a transformed version of the acquired runtime dataset~$\mathcal{R}$.
We describe the transformation of the runtime dataset in the subsequent section.
The features described in Section~\ref{sec:exdesign:data} serve as the input to the model.
Furthermore, it should be noted that a new prediction model is constructed in each iteration, as the runtime of the experiments (Line~5) dominates the training time of the model by a significant margin.
Finally, the score of the new solver, denoted by $\hat a$, is predicted with the prediction function~$f$ (Line~8).

@@

\paragraph{Uncertainty Sampling}

The uncertainty sampling strategy selects the instance that is closest to the model's decision boundary.
This is achieved by selecting the instance $e \in \mathcal{I} \setminus \tilde{\mathcal{I}}$ that minimizes $U(e)$, which is specified by the following equation.
%
\begin{equation*}
  \operatorname{U}(e) := \left\lvert \frac{1}{k} - \max_{n \in \left\lbrace 1, \dots, k \right\rbrace} f'\!\left(\hat{a}, e\right)_{n} \right\rvert
\end{equation*}

\paragraph{Information-Gain Sampling}

The information-gain sampling strategy selects the instance with the highest expected entropy reduction regarding the runtime labels of the instance.
To be more precise, we select the instance $e \in \mathcal{I} \setminus \tilde{\mathcal{I}}$ that maximizes $IG(e)$, which is specified in the following equation.
%
\begin{equation*}
  \operatorname{IG}(e) := \operatorname{H}(e) - \sum_{n = 1}^{k} f'(\hat{a}, e)_{n} \operatorname{\hat H}_n(e)
\end{equation*}
%
In the equation, $\operatorname{H}(e)$ denotes the entropy of the runtime labels $\gamma(a, e)$ over all $a \in \mathcal{A}$ and $\operatorname{H}(e, n)$ denotes the entropy of these labels plus $n$ as the runtime label for $\hat{a}$.
The term $\operatorname{\hat H}_n(e)$ is computed for every possible runtime label $n \in \{1, \dots, k\}$.
By maximizing the information gain, instances are selected that identify solvers with similar behavior.

@@

\paragraph{Wilcoxon Stopping Criterion}

The Wilcoxon stopping criterion is employed to halt the active-learning process based on the confidence that the predicted runtime labels of the new solver are sufficiently different from those of existing solvers.
This criterion is loosely inspired by Matricon et.~al.~\cite{MatriconAFSH21}.
The average $p$-value $W_{\hat{a}}$ of a Wilcoxon signed-rank test $\operatorname{w}(S,P)$, as given by the following equation, of the two runtime label distributions ${S=\{ \gamma(a, e) \mid e \in \mathcal{I} \}}$ for an existing solver $a$ and \mbox{$P=\{ f(\hat a, e) \mid e \in \mathcal{I} \}$} of the new solver $\hat{a}$ is used to assess the statistical significance of the predicted runtime labels.
%
\begin{equation*}
  W_{\hat{a}} := \frac{1}{\lvert \mathcal{A} \rvert} \sum_{a \in \mathcal{A}} \operatorname{w}(S, P)
\end{equation*}
%
In order to improve the stability of this criterion, an exponential moving average is employed to mitigate the impact of outliers, with the process terminated when the value of $W^{(i)}_{\exp}$ drops below a fixed threshold.
%
\begin{align*}
  W_{\exp}^{\left(0\right)} &:= 1\\
  W_{\exp}^{\left(i\right)} &:= \beta W_{\hat{a}} + \left(1 - \beta\right) W_{\exp}^{\left(i - 1\right)}
\end{align*}

\paragraph{Ranking Stopping Criterion}

In contrast, the ranking stopping criterion is less sophisticated than the previous one.
In this case, the active learning process is terminated if the ranking induced by the model's predictions (Equation~\ref{eq:rankingeq}) remains unchanged within the last~$l$~iterations.
However, it should be noted that the concrete values of the predicted score $s_{\hat a}$ may still change.
Nevertheless, in this case, the induced ranking is of sole interest.

@@

In light of the preceding instantiations of Algorithm~\ref{algALBenchmark}, this section outlines the experimental design employed in the presented study.
This includes an overview of the evaluation framework, the datasets employed, the hyperparameter selections, and the implementation details.

@@

As stated in the Introduction, this work addresses the \emph{New-Solver Problem} (cf.~Definition~\ref{def:new-solver-problem}).
As described in Section~\ref{sec:main:model}, a prediction model provides us with an estimated scoring value for the new solver~$\hat{a}$.
To evaluate the efficacy of a concrete instantiation of Algorithm~\ref{algALBenchmark}, which entails a specific choice for all subroutines, we perform cross-validation on our set of solvers.

This is demonstrated in Algorithm~\ref{alg:eval}.
Each solver assumes the role of the new solver~$\hat{a}$ once (Line~2) and is consequently excluded from the set of solvers $\mathcal{A}$ (Line~3).
After running our active-learning framework for solver~$\hat{a}$ (Line~4), we compute the value of both our optimization goals, i.e., ranking accuracy and runtime.
The \emph{ranking accuracy} $O_{\operatorname{acc}} \in \left[0, 1\right]$ (higher is better) is defined as the fraction of correctly ranked pairs of elements, namely, $\left(\hat{a}, a\right)$, regarding the ground-truth scoring function $\operatorname{par_2}$, where $a$ belongs to the set of all solvers $\mathcal{A}$ (Lines~5-8).
The \emph{fraction of runtime} required by the algorithm to reach its conclusion is represented by $O_{\operatorname{rt}}$, which lies within the interval $[0, 1]$ (lower is better).
This metric compares the accumulated runtimes over the sampled instances to the accumulated runtimes over all instances in the dataset (Lines~9-13).
Once all cross-validation results have been collected (Line~14), the output metrics are averaged (Line~15).

The procedure is designed to identify an approach that maximizes Equation~\eqref{eq:opt}, in which $\delta \in \left[0, 1\right]$ is used for linear weighting, with weighting factor $\delta \in \left[0, 1\right]$, between the two optimization goals $O_{\operatorname{acc}}$ and $O_{\operatorname{rt}}$.
The plotting of the approaches that maximize $O_\delta$ for all $\delta \in \left[0, 1\right]$ on an $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagram provides us with a Pareto front of the best approaches for different optimization-goal weightings.
%
\begin{equation}
  O_\delta := \delta O_{\operatorname{acc}} + \left(1 - \delta\right) \left(1 - O_{\operatorname{rt}}\right) \enspace \textrm{,}
  \label{eq:opt}
\end{equation} 
%

\subsection{Data}
\label{sec:exdesign:data}

In our experiments, we work with the dataset of the SAT~Competition~2022 Anniversary Track~\cite{sat2022}.
The dataset consists of 5355 instances with corresponding runtime data of 28 sequential SAT solvers.
We also use a database of 56 instance features\footnote{\url{https://benchmark-database.de/getdatabase/base_db}} from the Global Benchmark Database~(GBD) by Iser~et~al.~\cite{IserS18}.
They include instance size features and node distribution statistics for several graph representations of SAT instances, and are primarily inspired by the SATzilla~2012 features described in~\cite{features}.
All features are numeric and have no missing values.
We drop 10 of the 56 features due to zero variance.
In total, the prediction models have access to 46 instance features and 27 out of 28 runtime features, excluding the respective new solver~$\hat{a}$.

@@

\subparagraph{Ranking}

Regarding \emph{ranking} (cf.~Section~\ref{sec:main:model}), we experiment with the following instantiations.

@@

\subparagraph{Selection}

We experiment with the following instantiations of the instance \emph{selection} (cf. Section~\ref{sec:main:selection}).
Since the potential runtime of the experiments is orders of magnitude larger than the model update time, we consider incrementing our benchmark one instance at a time rather than using batches, which is also proposed in recent advances in active learning~\cite{SinhaED19,2019gaal}.
A drawback of this is the lack of parallel execution of runtime experiments.

@@

\subparagraph{Stopping}

We evaluate the following \emph{stopping} (cf. Section~\ref{sec:main:stopping}) criteria.

@@

\subsection{Runtime Discretization and Prediction}
\label{sec:exdesign:disc-pred}

One of the most important parts of our framework is the choice of solver runtime model to inform active learning selection decisions.
In this section, we detail how clustering solver runtimes can lead to nuanced runtime labels, and which machine learning models are best at predicting them.

@@

% MI: table with cluster prediction performance: (merged with existing text)
Since runtime prediction is quite challenging, we discretize the runtimes.
To define prediction targets for each instance, we cluster the solver runtimes into $k$ clusters.
We denote this adapted solver scoring function which assigns a solver to its cluster for each instance with $s_k$ as defined in Equation~\eqref{eq:rankingeq}.
In preliminary experiments, the most \emph{useful} labels were produced with hierarchical clustering and a log-single-link criterion.
In our chosen hierarchical procedure, each non-timeout runtime starts in a separate interval.
We then gradually merge intervals with the smallest single-link logarithmic distance until the desired number of partitions is reached.
The clustering is instance-specific, i.e., the bounds are chosen to produce groups of solvers with similar performance for each instance.
Other clustering approaches we have tried include hierarchical clustering with mean, median, and full-link criteria, as well as $k$-means and spectral clustering.

@@

Our final choice for the prediction model is the stacking ensemble of a quadratic discriminant analysis and a random forest.
Both of these models can learn non-linear relationships between the instance features and the runtime labels.
Fig.~\ref{fig:confusion-matrices} shows the confusion matrices of the stacking ensemble for predicting timeouts and $C_3$ clusters.
Both perform reasonably well, with the $C_3$ cluster prediction model performing slightly worse than the timeout prediction model.
Notably, for the $C_3$ prediction model, there is less confusion between the non-timeout labels (labels 1 and 2) and the timeout label (label 3) than between the non-timeout labels (labels 1 and 2).
This suggests that deciding on timeouts is easier than distinguishing between different groups of solvers.
